---
title: "Clustering"
output: 
  ioslides_presentation
  #html_document: default
---

```{r setup, include=FALSE}
library(datasets)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```

## Cluster analysis

Grouping objects so that objects in the same group are more similar to each other than to objects of other groups.

``` {r cars}
ggplot(cars, aes(x=speed, y=dist)) + geom_point(color="black")
```

## Cluster analysis

Grouping objects so that objects in the same group are more similar to each other than to objects of other groups.

``` {r cars-clustering}
clusters <- kmeans(cars$speed * cars$dist, 3, iter.max = 50)
clustered <- data.frame(
  speed=cars$speed, 
  dist=cars$dist, 
  color=ifelse(clusters$cluster==1, "red", ifelse(clusters$cluster==2, "blue", "green")))
                   
ggplot(clustered, aes(x=speed, y=dist)) + geom_point(aes(color=color)) + theme(legend.position="none")
```

## Agglomerative Hierarchical Clustering

Clusters: 2  
Input: 2, 8, 4, 11, 3

``` {r ahc1, fig.width=8, fig.height=1}

gg_circle <- function(r, xc, yc, color="black", fill=NA, ...) {
    x <- xc + r*cos(seq(0, pi, length.out=100))
    ymax <- yc + r*sin(seq(0, pi, length.out=100))
    ymin <- yc + r*sin(seq(0, -pi, length.out=100))
    annotate("ribbon", x=x, ymin=ymin, ymax=ymax, color=color, fill=fill, ...)
}

gg_rect <- function(r, x, color) {
  annotate("rect", xmin=x-r-0.05, xmax=x+r+0.05, ymin=0.2, ymax=1.8, alpha=.2, fill=color)
}

original <- data.frame(x=c(2, 3, 4, 8, 11), y=c(1,1,1,1,1), color=c(0,0,0,0,0))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=c(1:11)) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") 
  
```

## Agglomerative Hierarchical Clustering

Clusters: 2  
Input: 2, 8, 4, 11, 3

``` {r ahc2, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11), y=c(1,1,1,1,1), color=c(1,1,0,0,0))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") + 
  gg_rect(r=.5, x=2.5, color="skyblue")
```

## Agglomerative Hierarchical Clustering

Clusters: 2  
Input: 2, 8, 4, 11, 3

``` {r ahc3, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11), y=c(1,1,1,1,1), color=c(1,1,0,0,0))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") + 
  gg_rect(r=.5, x=2.5, color="skyblue")
```

Linkage:

```
single([2,3], [4]) = 1
complete([2,3], [4]) = 2
centroid([2,3], [4]) = 1.5
```

## Agglomerative Hierarchical Clustering

Clusters: 2  
Input: 2, 8, 4, 11, 3


``` {r ahc4, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11), y=c(1,1,1,1,1), color=c(1,1,1,0,0))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") + 
  gg_rect(r=1, x=3, color="skyblue")
```

Linkage:

```
single([2,3], [4]) = 1
complete([2,3], [4]) = 2
centroid([2,3], [4]) = 1.5
```

## Agglomerative Hierarchical Clustering

Clusters: 2  
Input: 2, 8, 4, 11, 3

``` {r ahc5, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11), y=c(1,1,1,1,1), color=c(1,1,1,0,0))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") + 
  gg_rect(r=1, x=3, color="skyblue")
```

Linkage:

```
single([2,3,4], [8]) = 4, single([8], [11]) = 3
complete([2,3,4], [8]) = 6, complete([8], [11]) = 3
centroid([2,3,4], [8]) = 5, centroid([8], [11]) = 3
```

## Agglomerative Hierarchical Clustering

Clusters: 2  
Input: 2, 8, 4, 11, 3

``` {r ahc6, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11), y=c(1,1,1,1,1), color=c(1,1,1,0,0))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") + 
  gg_rect(r=1, x=3, color="skyblue") +
  gg_rect(r=1.5, x=9.5, color="black")
```

Linkage:

```
single([2,3,4], [8]) = 4, single([8], [11]) = 3
complete([2,3,4], [8]) = 6, complete([8], [11]) = 3
centroid([2,3,4], [8]) = 5, centroid([8], [11]) = 3
```

## K-means

Clusters: 2  
Input: 2, 8, 4, 11, 2

``` {r k1, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11), y=c(1,1,1,1,1), color=c(0,0,0,0,0))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none")
```

## K-means

Clusters: 2  
Input: 2, 8, 4, 11, 2

``` {r k2, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11, 2, 4), y=c(1,1,1,1,1,1.25,1.25), color=c(0,0,0,0,0,1,2))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none")
```

Initial centroids: A=2, B=4

## K-means

Clusters: 2  
Input: 2, 8, 4, 11, 2

``` {r k3, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11, 2, 4), y=c(1,1,1,1,1,1.25,1.25), color=c(0,0,0,0,0,1,2))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") +
  gg_rect(r=.5, x=2.5, color="skyblue") +
  gg_rect(r=3.5, x=7.5, color="black")
```

Initial centroids: A=2, B=4

## K-means

Clusters: 2  
Input: 2, 8, 4, 11, 2

``` {r k4, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11, 2.5, 7.5), y=c(1,1,1,1,1,1.25,1.25), color=c(0,0,0,0,0,1,2))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") +
  gg_rect(r=.5, x=2.5, color="skyblue") +
  gg_rect(r=3.5, x=7.5, color="black")
```

New centroids: A=2.5, B=7.5

## K-means

Clusters: 2  
Input: 2, 8, 4, 11, 2

``` {r k5, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11, 2.5, 7.5), y=c(1,1,1,1,1,1.25,1.25), color=c(0,0,0,0,0,1,2))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") +
  gg_rect(r=1, x=3, color="skyblue") +
  gg_rect(r=1.5, x=9.5, color="black")
```

New centroids: A=2.5, B=7.5


## K-means

Clusters: 2  
Input: 2, 8, 4, 11, 2

``` {r k6, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11, 3, 9.5), y=c(1,1,1,1,1,1.25,1.25), color=c(0,0,0,0,0,1,2))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") +
  gg_rect(r=1, x=3, color="skyblue") +
  gg_rect(r=1.5, x=9.5, color="black")
```

## K-means

Clusters: 2  
Input: 2, 8, 4, 11, 2

``` {r k7, fig.width=8, fig.height=1}

original <- data.frame(x=c(2, 3, 4, 8, 11, 3, 9.5), y=c(1,1,1,1,1,1.25,1.25), color=c(0,0,0,0,0,1,2))

ggplot(original, aes(x,y,colour=color)) + geom_point() + 
  ylim(0, 2) +
  scale_x_discrete(limits=1:11) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.background = element_blank(),
        legend.position="none") +
  gg_rect(r=1, x=3, color="skyblue") +
  gg_rect(r=1.5, x=9.5, color="black")
```

Centroids/membership is stable

## Clustering in two dimensions

**Euclidian distance**  

$d_\epsilon(X,Y) = \sqrt{\big|X_0 - Y_0\big|^2 + \big|X_1 - Y_1\big|^2 + ... + \big|X_n - Y_n\big|^2}$


**p-norm**

$d_p(X, Y) = \bigg(\big|X_0 - Y_0\big|^2 + \big|X_1 - Y_1\big|^2 + ... + \big|X_n - Y_n\big|^2\bigg)^{1/p}$

- Manhattan ($p=1$)
- Euclidian ($p=2$)
-Chebyshev ($p=∞$)


Other: **Levenstein-Damerau** (strings), graph distances
    
